Dissecting the relationship between protein structure and sequence variation

### 2015-05-19 manuscript/voronoi/mnv.tex
MN update

    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure. An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}. In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}. Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN. A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $5\AA$ to $18\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014}. \\

@@

        \begin{equation}
            \label{eqn:wcn_pwrl}
            \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}}.
        \end{equation}

@@

@@

    There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \cite{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}). \\

@@

>>>>

    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure. An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}. In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009},

    \begin{equation}
        \label{eqn:cn}
        \text{CN}_{i} = \sum^N_{j\neq i} H\big(r_0-r_{ij}\big),
    \end{equation}

    \noindent in which $r_{ij}$ represents the distance between sites $i$ \& $j$ and,

    \begin{equation}
        \label{eqn:heaviside}
        H\big(r_0-r_{ij}\big) = \int^{r_0-r_{ij}}_{-\infty}\delta\big(x)\operatorname{x},
    \end{equation}

    \noindent is the Heaviside step function, with $\delta(x)$ standing for the Dirac delta function. Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN. A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r_0$ in the definition of CN. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $5\AA$ to $18\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014}. \\

@@

    \begin{equation}
        \label{eqn:wcn_pwrl}
        \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{\alpha=2}}.
    \end{equation}

@@

    \subsection*{Contact Number Definition and Long-Range Amino Acid Interactions}
    \label{sec:cnwcn}

    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigated the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and amino acid fluctuations (B factor), for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns. \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).

    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcn}.

@@

    There is already an extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}). \\

@@

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{}
        \label{fig:wcnh}
    \end{figure}

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN$_\operatorname{P}$) with power-law kernel as given by Eqn. \ref{eqn:wcn_pwrl}, with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$. The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution. The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.  Note that for $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ for $\alpha<0$. In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot. The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
        \label{fig:wcnp}
    \end{figure}
